Variance reduction for Markov chains via martingale representations

### 2020-05-08 aap/Revision/vr-mcmc-revision-final.tex
update

Let $f$ be a bounded Borel function and $(X^x_{p})_{p \in \nset_0}$ be a Markov chain generated by ULA with constant step size $\gamma$.
Let $c_0>0$. Then, there exist $C < \infty$ and $\gamma_0 > 0$, such that for all $n \in \nset$, $\gamma \in (0,\gamma_0]$ satisfying $n\gamma > c_0$, and $x \in \rset^d$,

@@

\PVar\left[\pi_n^N(f)\right]\leq C (n \gamma)^{-1} V(x) \eqsp.

@@

Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ is $V-$geometrically ergodic in the sense of \eqref{eq:v_ergodicity} with $V(x) = 1+|x|^2$ and $\rho\in (0,1) $ specified in Lemma~\ref{lem:v_ergodicity}. Due to Lemma~\ref{lem:var_bound},
\[
\PVar[\pi_n^N(f)]\lesssim \frac{1}{n(1-\rho^{1/2})} + \left(\frac{\rho^NV(x)+1}{n(1-\rho^{1/2})}\right)^2
\]
It follows from \eqref{eq:rho-ula} that $1-\rho^{1/2} \geq C\gamma$ with a constant $C$ not depending on $\gamma,N$ and $n$, thus yielding an estimate
\[
\PVar[\pi_n^N(f)] \lesssim \frac{1}{n\gamma} + \left(\frac{\rho^NV(x)+1}{n\gamma}\right)^2.
\]

@@

\begin{definition}
\label{def:v_ergodicity}
Let $(X_p)_{p \in \nset_0}$ be a Markov chain taking values in some space $X$ with the Markov kernel $Q$ and stationary distribution $\pi$. We say that $(X_p)_{p \in \nset_0}$ is $V-$geometrically ergodic for a given function $V: X \rightarrow [1; +\infty)$ if there exist real numbers $C > 0$ and $0 < \rho < 1$ such that for any $n \in \nset$,
\begin{equation}
\label{eq:v_ergodicity}
d_V(\delta_xQ^{n},\pi) \leq C\rho^n V(x)
\end{equation}
\end{definition}
In  the sequel, without loss of generality, we assume that $(X_p)_{p \in \nset_0}$ follows ULA dynamics \eqref{eq:chain} with $\nabla U(0)=0$ and  set
\begin{equation}
\label{eq:definition-V}
V(x) = 1 + |x|^2 \eqsp.
\end{equation}
\begin{lem}
  \label{lem:quadratic_behaviour}
  Assume {\bf (H1)} and {\bf (H2)}. Then there exists $K_2 \geq 0$ such that for any $ | x | \geq K_2$ it holds $\ps{\nabla U(x)}{x} \geq (\mathfrak{m}/2) | x |^2$.
\end{lem}
\begin{proof}
 Suppose that the function $\Delta$ in {\bf (H2)} is supported in the ball $\{x:|x|\leq K_1\}$, then we have 
  \begin{align*}
    \ps{\nabla U(x)}{x}
    &= \int_{0}^{K_1/| x |} D^2 U(t x ) [x^{\otimes 2}] \rmd t + \int_{K_1/| x |} ^ 1 D^2 U(t x ) [x^{\otimes 2}] \rmd t\\
    & \geq \mathfrak{m} |x |^2 \{1- K_1 (1 +L/\mathfrak{m})   / | x | \}\eqsp,\quad \mbox{for all} \quad  | x |\geq K_1,
  \end{align*}
which proves the  statement with $K_2=2K_{1}(1+L/\mathfrak{m}).$
\end{proof}
\begin{lem}
\label{lem:drift}
 Assume {\bf (H1)} and {\bf (H2)}.
 Then, for any $\gamma \in (0,\bar{\gamma}]$ with $\overline{\gamma} = \mathfrak{m}/2L^2$  the kernel $Q_\gamma$ from \eqref{eq:ula_kernel} satisfies drift condition
 \begin{equation}
\label{eq:lyapunov}
Q_{\gamma}V(x) \leq \rho^{\gamma} V(x) + \gamma C
\end{equation}
with   $\rho = \exp{\left(-\frac{\mathfrak{m}}{2}\right)}$, $C = 2K_2^2(L+\mathfrak{m}) + d + \mathfrak{m}$ with $K_2$ from Lemma~\ref{lem:quadratic_behaviour}, and $\mathfrak{m}$ from {\bf (H2)}.
\end{lem}

@@

\begin{lem}
\label{lem:v_ergodicity}
 Assume  {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}= \min(\mathfrak{m} / 4L^2,1/L)$, and for any $x \in \rset^d$  it holds
$$
d_V(\delta_xQ_{\gamma}^{n},\pi_\gamma) \leq C\rho^{ n}\left(V(x) + \pi_\gamma(V)\right)
$$
with $V(x) = 1 + |x|^2$ and constants
$$
C = \left(1 + \exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} - \frac{2b}{2+K^2})}\right),
$$

$$
b = \gamma(2K_2^2(L+\mathfrak{m}) + d + \mathfrak{m}), \quad \overline{b} = b\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} + K^2 + 1, 
$$

\begin{eqnarray*}
\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right),
\end{eqnarray*}

\begin{eqnarray}
\label{eq:rho-ula}
\rho = \exp\left\{-\frac{\gamma m}{4}\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} + \frac{2b}{K^2+2})}}\right\},
\end{eqnarray}
where $K_2 > 0$ is from Lemma~\ref{lem:quadratic_behaviour} and $K>0$ is a positive parameter such that 
$$
K^2 > 4\frac{2K_2^2(L+m) + d + m}{m}.
$$
\end{lem}

\begin{proof} Note that the condition {\bf (H1)} implies that for any $x,y \in \rset^d$,
\[
|x-y-\gamma(\nabla U(x) - \nabla U(y))|^2 \leq (1 + 2L\gamma + \gamma^2)|x-y|^2
\] 
Hence,   \cite[Corollary 5]{DurmusBortoli} implies that
\[
\|\delta_xQ^{[1/\gamma]}_{\gamma} - \delta_yQ^{[1/\gamma]}_{\gamma}\|_{\text{TV}} \leq 2(1-\varepsilon)
\]
with $\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right)$ for all $x,y$ such that $|x|,|y| \leq K$, which means that the kernel $Q_{\gamma}$ satisfies $([1/\gamma],\varepsilon)$-Doeblin condition on the set $\{x: V(x) < 1+K^2\}$. Together with drift condition \eqref{eq:lyapunov} it allows to apply \cite[Theorem~19.4.1]{moulines2018} with appropriate constants.
\end{proof}

>>>>

Let $f$ be a bounded Borel function and $(X^x_{p})_{p \in \nset_0}$ be a Markov chain generated by ULA with constant step size $\gamma$. Then, there exist $C < \infty$ and $\gamma_0 > 0$, such that for all $n \in \nset$, $\gamma \in (0,\gamma_0]$ and $x \in \rset^d$,

@@

\PVar\left[\pi_n^N(f)\right]\leq C\left(\frac{1}{n\gamma} + \left(\frac{V(x)+1}{n\gamma}\right)^2\right) \eqsp.

@@

Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ generated by ULA  is $V-$geometrically ergodic in the sense of \eqref{eq:v_ergodicity} with $V(x) = 1+|x|^2$ and $\rho\in (0,1) $ due to Lemma~\ref{lem:v_ergodicity}. It follows from Lemma~\ref{lem:var_bound}, that
\[
\PVar[\pi_n^N(f)]\lesssim \frac{1}{n(1-\rho^{1/2})} + \left(\frac{\rho^NV(x)+1}{n(1-\rho^{1/2})}\right)^2
\]
It follows from \eqref{eq:rho-ula} that $1-\rho^{1/2} \geq C\gamma$ with a constant $C$ not depending on $\gamma,N$ and $n$, thus yielding an estimate
\[
\PVar[\pi_n^N(f)] \lesssim \frac{1}{n\gamma} + \left(\frac{\rho^NV(x)+1}{n\gamma}\right)^2.
\]

@@

\begin{definition}
\label{def:v_ergodicity}
Let $(X_p)_{p \in \nset_0}$ be a Markov chain taking values in some space $X$ with the Markov kernel $Q$ and stationary distribution $\pi$. We say that $(X_p)_{p \in \nset_0}$ is $V-$geometrically ergodic for a given function $V: X \rightarrow [1; +\infty)$ if there exist real numbers $C > 0$ and $0 < \rho < 1$ such that for any $n \in \nset$,
\begin{equation}
\label{eq:v_ergodicity}
d_V(\delta_xQ^{n},\pi) \leq C\rho^n V(x)
\end{equation}
\end{definition}
In  the sequel, without loss of generality, we assume that $(X_p)_{p \in \nset_0}$ follows ULA dynamics \eqref{eq:chain} with $\nabla U(0)=0$ and  set
\begin{equation}
\label{eq:definition-V}
V(x) = 1 + |x|^2 \eqsp.
\end{equation}
\begin{lem}
  \label{lem:quadratic_behaviour}
  Assume {\bf (H1)} and {\bf (H2)}. Then there exists $K_2 \geq 0$ such that for any $ | x | \geq K_2$ it holds $\ps{\nabla U(x)}{x} \geq (\mathfrak{m}/2) | x |^2$.
\end{lem}
\begin{proof}
 Suppose that the function $\Delta$ in {\bf (H2)} is supported in the ball $\{x:|x|\leq K_1\}$, then we have 
  \begin{align*}
    \ps{\nabla U(x)}{x}
    &= \int_{0}^{K_1/| x |} D^2 U(t x ) [x^{\otimes 2}] \rmd t + \int_{K_1/| x |} ^ 1 D^2 U(t x ) [x^{\otimes 2}] \rmd t\\
    & \geq \mathfrak{m} |x |^2 \{1- K_1 (1 +L/\mathfrak{m})   / | x | \}\eqsp,\quad \mbox{for all} \quad  | x |\geq K_1,
  \end{align*}
which proves the  statement with $K_2=2K_{1}(1+L/\mathfrak{m}).$
\end{proof}
\begin{lem}
\label{lem:drift}
 Assume {\bf (H1)} and {\bf (H2)}.
 Then, for any $\gamma \in (0,\bar{\gamma}]$ with $\overline{\gamma} = \mathfrak{m}/2L^2$  the kernel $Q_\gamma$ from \eqref{eq:ula_kernel} satisfies drift condition
 \begin{equation}
\label{eq:lyapunov}
Q_{\gamma}V(x) \leq \lambda^{\gamma} V(x) + \gamma b
\end{equation}
with   $\lambda = \exp{\left(-\frac{\mathfrak{m}}{2}\right)}$, $b = 2K_2^2(L+\mathfrak{m}) + d + \mathfrak{m}$ with $K_2$ from Lemma~\ref{lem:quadratic_behaviour}, and $\mathfrak{m}$ from {\bf (H2)}.
\end{lem}

@@

\begin{lem}
\label{lem:v_ergodicity}
 Assume  {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}= \min(\mathfrak{m} / 2L^2,1/L)$, and for any $x \in \rset^d$  it holds
$$
d_V(\delta_xQ_{\gamma}^{n},\pi_\gamma) \leq C\rho^{\gamma n} V(x)
$$
with $V(x) = 1 + |x|^2$ and some $0 < \rho < 1$ and $C > 0$.
%$$
%C = \left(1 + \exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} - \frac{2b}{2+K^2})}\right),
%$$
%
%
%
%$$
%b = \gamma(2K_2^2(L+\mathfrak{m}) + d + \mathfrak{m}), \quad \overline{b} = b\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} + K^2 + 1, 
%$$
%
%\begin{eqnarray*}
%\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right),
%\end{eqnarray*}
%
%\begin{eqnarray}
%\label{eq:rho-ula}
%\rho = \exp\left\{-\frac{\gamma m}{4}\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} + \frac{2b}{K^2+2})}}\right\},
%\end{eqnarray}
%where $K_2 > 0$ is from Lemma~\ref{lem:quadratic_behaviour} and $K>0$ is a positive parameter such that 
%$$
%K^2 > 4\frac{2K_2^2(L+m) + d + m}{m}.
%$$
\end{lem}

\begin{proof} Note that the condition {\bf (H1)} implies that for any $x,y \in \rset^d$,
\[
|x-y-\gamma(\nabla U(x) - \nabla U(y))|^2 \leq (1 + 2L\gamma + \gamma^2)|x-y|^2
\] 
Hence,   \cite[Corollary 5]{DurmusBortoli} implies that for all $x,y$ such that $|x|,|y| \leq K$,
\[
\|\delta_xQ^{\lceil 1/\gamma \rceil}_{\gamma} - \delta_yQ^{ \lceil 1/\gamma \rceil}_{\gamma}\|_{\text{TV}} \leq 2(1-\varepsilon)
\]
where $\lceil \cdot \rceil$ denotes an upper integer part and $\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right)$. Note also that the drift condition \eqref{eq:lyapunov} implies that
\[
Q^{[1/\gamma]}_{\gamma}V(x) \leq \lambda V(x) + \frac{\gamma b(1-\lambda^{\gamma(\lceil 1/\gamma\rceil + 1)})}{1-\lambda^{\gamma}} \leq \lambda V(x) + \frac{b}{\lambda \log{1 /\lambda}} 
\]
where the last inequality is due to \cite[Lemma 1]{durmus:moulines:2017}. Hence, the kernel $Q^{\lceil 1/\gamma \rceil}_{\gamma}$ satisfies $(1,\varepsilon)$-Doeblin condition on the set $\{x: V(x) < 1+K^2\}$ and drift condition \eqref{eq:lyapunov} with appropriate constants. Now the statement of the lemma follows from \cite[Theorem~19.4.1]{moulines2018}.
\end{proof}
